---
title: "Problem set 1"
author: "BALYA SHUKLA"
date: "02/11/2019"
---

# Details

- Who did you collaborate with: Self and the internet. 
- Approximately how much time did you spend on this problem set: ~4 hours.
- What, if anything, gave you the most trouble: I struggled with figuring out why Oceania did not show up on the violin chart. I don't think I have the right answer. 


# Task 1

I watched the short video and completed the series of R Markdown tutorials.


# Task 2

## Life expectancy in 2007

```{r load-packages, warning=FALSE, message=FALSE}
# Notice the warning=FALSE and message=FALSE in the chunk options. R spits out 
# a lot of messages when you load tidyverse and we don't want those in the 
# final document.

library(tidyverse)  # This loads ggplot2, dplyr, and other packages you'll need
library(gapminder)  # This loads the gapminder data
```

Let's first look at the first few rows of data:

```{r view-data}
head(gapminder)
```

Right now, the `gapminder` data frame contains rows for all years for all countries. We want to only look at 2007, so we create a new data frame that filters only rows for 2007.

Note how there's a weird sequence of characters: `%>%`. This is called a *pipe* and lets you chain functions together. We could have also written this as `gapminder_2007 <- filter(gapminder, year == 2007)`.

```{r filter-2007}
gapminder_2007 <- gapminder %>%
  filter(year == 2007)

head(gapminder_2007)
```

Now we can plot a histogram of 2007 life expectancies with the default settings:

```{r plot-2007-1}
ggplot(data = gapminder_2007, 
       mapping = aes(x = lifeExp)) +
  geom_histogram()
```

R will use 30 histogram bins by default, but that's not always appropriate, and it will yell at you for doing so. **Adjust the number of bins to 2, then 40, then 100.** **What's a good number for this data? Why?**

TYPE YOUR ANSWER HERE.I would pick bins = 40 because it clearly comunicates the information. You can see that the data is skewed to the left where majority of the coutries had the life expectancy of 60 or greater in 2017 with some outliers with life expectancy greater than 85 and less that 45. 100 bins might give more detailed information, but it is not as visually pleasing as 40; 40 does the same job as 100 but is more visually pleasing. And the reason I wouldn't pick 2 is because it lacks details and tells the viewer very little about about the life expectancy in 2017. 

```{r plot-2007-2}
ggplot(data = gapminder_2007, 
       mapping = aes(x = lifeExp)) +
  geom_histogram(bins = 40)
```

## Average life expectancy in 2007 by continent

We're also interested in the differences of life expectancy across continents. First, we can group all rows by continent and calculate the mean

This is where the `%>%` function is actually super useful. Remember that it lets you chain functions together—this means we can read these commands as a set of instructions: take the `gapminder` data frame, filter it, group it by continent, and summarize each group by calculating the mean. Without using the `%>%`, we could write this same chain like this: `summarize(group_by(filter(gapminder, year == 2007), continent), avg_life_exp = mean(lifeExp))`. But that's *awful* and impossible to read and full of parentheses that can easily be mismatched.

```{r calc-mean}
gapminder_cont_2007 <- gapminder %>%
  filter(year == 2007) %>% 
  group_by(continent) %>%
  summarize(avg_life_exp = mean(lifeExp))

head(gapminder_cont_2007)
```

Let's plot these averages as a bar chart:

```{r plot-2007-bar}
ggplot(data = gapminder_cont_2007, 
       mapping = aes(x = continent, y = avg_life_exp, fill = continent)) + 
  geom_col()
```

Then, let's plot them as density distributions. We don't need to use the summarized data frame for this, just the original filtered `gapminder_2007` data frame:

```{r plot-2007-density}
ggplot(data = gapminder_2007, 
       mapping = aes(x = lifeExp, fill = continent)) + 
  geom_density()
```

Now let's plot life expectancies as violin charts. These are the density distributions turned sideways:

```{r plot-2007-violin}
ggplot(data = gapminder_2007, 
       mapping = aes(x = continent, y = lifeExp, fill = continent)) + 
  geom_violin()
```

Finally, we can add actual points of data for each country to the violin chart:

```{r plot-2007-violin-points}
ggplot(data = gapminder_2007, 
       mapping = aes(x = continent, y = lifeExp, fill = continent)) + 
  geom_violin() +
  geom_point()
```

The bar chart, density plot, violin plot, and violin plot + points each show different ways of looking at a single number—the average life expectancy in each continent. **Answer these questions:**

- Which plot is most helpful? I would say bar chart because it shows the average life expectancy for each continent. It is easy to read and understand. Removing the legend would reduce the redundancy of the chart, and if we rearranged the bars in ascending or descending order, it will tell a little more about the average life expectancy of each continent. But then again, mean is not a good measure to judge the entire population of a continent because some countries could be outliers and that can affect the mean. In this case, I would say the voilin plot is more insightful expect it does not show the density distribution of Oceania. 
- Which ones show variability? TYPE YOUR ANSWER HERE.I want to say voilin chart because it clearly shows the density distribution. However, it does not include Ocenia. So I am leaning more towards density distribution curve. But my main concern with it is the readabilty because the color fill is overlapping with one another, and it is hard to pin point the highest peak of each density curve. Also, the voilin curve with points, does not exactly add value to the voilin graph according to me. That is because each point is a country and unless we know what each point represents, it is of no use (except for Oceania).
- What's going on with Oceania? TYPE YOUR ANSWER HERE.In an article, Violin Plots 101: Visualizing Distribution and Probability Density, by JOEL CARRON, Carron mentions that a violin plot is a combination of box plot and kernel density plot (probability density chart) and it shows peaks in the data. It consists of the median, interquartile range, and 95% confidence level. It could be that the two countries in Oceania are very close to the median and which is why there is no density distribution of the continent. 


# Task 3: R and ggplot

```{r load-libraries-1, warning=FALSE, message=FALSE}
# Technically you don't need to run this because we loaded tidyverse earlier in Task 2
library(tidyverse)
```

## 3.2.4

**1: Run `ggplot(data = mpg)`. What do you see?**

```{r blank-plot, fig.width=2, fig.height=2}
ggplot(data = mpg)
```

TYPE YOUR ANSWER HERE. (hint: I gave you the answer in the chunk name) A blank ggplot with a width and height of 2 inches. 

(Notice how I used `fig.width` and `fig.height` in the chunk options. You can click on the little gear icon in the far left of the chunk to change other options.)

**2: How many rows are in `mpg`? How many columns?**

There are 234 rows and 11 columns in `mpg`.
```{r mpg-details}
nrow(mpg)
ncol(mpg)

# Or
dim(mpg)

# Or
mpg
```

**3: What does the `drv` variable describe? Read the help for `?mpg` to find out.**

```{r}
?mpg
```


TYPE YOUR ANSWER HERE.
drv
f = front-wheel drive, r = rear wheel drive, 4 = 4wd

**4: Make a scatterplot of `hwy` vs `cyl`.**

```{r hwy-cyl-scatterplot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = hwy))
```

**5: What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?**

```{r class-drv-scatterplot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))
```

TYPE YOUR ANSWER HERE.
The class variable is the type of car and the drv variable is the type of wheel drive. They are both categorical variables and plotting them in an x and y axis does not give any valuable information expect for how different car types vary in the type of wheel drive. There could be points overlapping, and we can't tell the count from this graph alone. 

## 3.3.1

**1: What’s gone wrong with this code? Why are the points not blue?**
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```
ANSWER: 
The points are not blue because the color = "blue" is inside the aesthetics argument which adds another parameter called "blue". The color will not go under the asthetics because it is not conveying any information about the variable; it is only changing the appearance of the plot and should go outside the asthetics argument. The correct code would be: 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```


**2: Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?**


ANSWER: 
```{r}
?mpg
mpg
```
When mpg is run, you can see that each column header also shows the variable type. For example, under manufacturer, it says <chr>, which means it is a character variable. There are 6 character variables, 4 integers, and 1 double floating point number. 

**3: Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?**

ANSWER: 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = cyl, color = year, shape = cty))
```
A continuous variable can not be mapped to shape. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = cyl, color = year))
```
Using size works better for continuous variables and should not be used for discrete variables because size is an ordered aesthetic and categorical variables are unordered. 

Color works better if using categorical variables because for continuous variables the color is displayed in a color grid from the darkest to the lightest. It is hard to differentiate between different years in this context. 

**4: What happens if you map the same variable to multiple aesthetics??**

ANSWER:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = class, color = class))
```
It doesn't add or reduce value to the graph. We can take the legend out in this case. 


**5: What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)?**

ANSWER: 

```{r}
?geom_point
```
The stroke aesthetic can be used to modify the width of the border for shapes that have a border, for example: shape 21 which is a circle with a fill and a border. 

## 3.5.1

**1: What happens if you facet on a continuous variable?**

ANSWER:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ cyl)
```
If facet is used for a continous variable, like categorical variable, it will create a new chart for each value. In this case, there are only 4 unique values, but if we were to use facet on cty, we would get several charts for each value and that can be cumbersome. 

**2: What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?**
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
```

ANSWER:

The empty cells in plot with facet_grid(drv ~ cyl) mean that there are no observation that fall under rear wheel drive and 4 cylinders. Like the point plot shows above that there are no observation that have rear wheel drive and 4 cylinders in the dataset. 


**3: What plots does the following code make? What does . do?**
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

ANSWER:

Placing "." after "~" (in place of a variable), will put the variable placed before ~ in rows. Placing "." before "~" (in place of a variable), will put the variable placed after ~ in columns. You would want to use "." to not facet in the rows or columns.  


**4: Take the first faceted plot in this section:                                                  What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?**
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

ANSWER:

Using faceting makes the comparison of each category easier. You can see the general hwy and displ of each category. With color, it can become hard to compare when we have 7 or more different colors. However, it can be a little tedious to isolate individual points with faceting. 


## 3.6.1

**1: What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?**

Line chart: 
  geom_line()
Boxplot: 
  geom_boxplot()
Histogram: 
  geom_histogram() >> uses 30 bins as a default, but can change the number of bins by using this code:
  geom_histogram(bins = 40)
Area chart: 
  geom_area()

**2: Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.**

ANSWER: 

This code can be broken down into 4 parts:
data = mpg > calls the mpg dataset.
mapping = aes(x = displ, y = hwy, color = drv)) >> the mapping function will be treated as the global mapping that will be apply to each geom in the graph. The aesthetic argument calls x-axis to be engine displacement, in litres, y-axis to be highway miles per gallon, and color of the points to be wheel drive type. 
geom_point() >> tells the plot type which is point plot. 
geom_smooth(se = FALSE) >> geom_smooth is the smooth line fitted to the data. The se argument displays confidence interval around smooth which is set as TRUE by default. If changed to FALSE, the shaded part around the smoothed line will disappear. 

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```



**3: What does show.legend = FALSE do? What happens if you remove it? Why do you think I used it earlier in the chapter?**

ANSWER:

show.legend = FALSE will remove the legend. If you remove it, the legend will show. I am guessing it was used earlier in the chapter to show that the legend will show by default and to hide the legend, we can use the argument show.legend = FALSE. 
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class), show.legend = FALSE) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
   
```


**4: What does the se argument to geom_smooth() do?**

ANSWER:

The se argument displays confidence interval around smooth which is set as TRUE by default. If changed to FALSE, the shaded part around the smoothed line will disappear. 

**5: Will these two graphs look different? Why/why not?**

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

ANSWER: 

They will look the same because the first one has global mapping applicable to all geoms in the ggpplot where as the second one does individual mapping for each plot type. 


## 3.8.1

**1: What is the problem with this plot? How could you improve it?**

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```

ANSWER: 

Overplotting is the problem because many points overlap each other. To fix the problem we can use the following code: 

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point(position = "jitter", alpha = 1/5)
```


**2: What parameters to geom_jitter() control the amount of jittering?**

ANSWER: 

```{r}
?geom_jitter
```

Width and height parameters control the amount of vertical and horizontal jitter.

